---
title: "EM and contrast demo"
author: "Kim Johnson"
date: "11/5/2020"
output: html_document
---

# Load packages and libraries
```{r, eval=FALSE}
#install.packages("survminer") 
#install.packages("ggfortify")
#install.packages("haven")
#install.packages("multcomp")
library(tidyverse) #for various packages
library(haven) #for read_csv
library(multcomp)
library(lmtest)

#load data for this exercise

cerv<- read_csv("SEERcerv2004to2015.csv") #load data for this exercise
```

```{r}
#provide shorter names for variables
names(cerv)<-c("ID", "age_category","yr_dx", "sex", "race",
               "insurance", "marital", "%pov", "%edu", "cause_spec_death", 
               "first_primary", "surv_mo", "vital_stat", "Stage") 


#Recode cause specific death as 1/0 if the person died/did not die of cervical cancer 
cerv$event_f[
    cerv$cause_spec_death=="Dead (attributable to this cancer dx)"]<-1
cerv$event_f[
    cerv$cause_spec_death=="Alive or dead of other cause"|
    cerv$cause_spec_death=="N/A not first tumor"]<-0

#Recode insurance status as a factor variable and label it
cerv$insurance_f[
  cerv$insurance=="Insured"]<-0
cerv$insurance_f[
 cerv$insurance=="Any Medicaid"]<-1
cerv$insurance_f[
  cerv$insurance=="Uninsured"]<-2
cerv$insurance_f<-factor(cerv$insurance_f,
      levels = c(0,1,2),
      labels = c("Insured", "Any Medicaid", "Uninsured"))

#Recode race
cerv$race[
  cerv$race=="Non-Hispanic White"]<-0
cerv$race[
 cerv$race=="Non-Hispanic Black"]<-1
cerv$race[
 cerv$race=="Hispanic (All Races)"]<-2
cerv$race<-factor(cerv$race,
      levels = c(0,1,2),
      labels = c("White", "Black", "Hispanic"))

cerv$surv_mo<-as.numeric(cerv$surv_mo)
```

# Complete case analysis exclusions, exclude missing data
```{r}
cerv<-cerv[which(!is.na(cerv$race) & !is.na(cerv$insurance_f) & !is.na(cerv$event_f) & !is.na(cerv$surv_mo)),]
```

# Run Cox models to test the hypothesis that race modifies the association between insurance and the hazard of death.
```{r}
#Model with insurance and race
insurance_mod_race<-coxph(Surv(surv_mo, event_f)~insurance_f + race, cerv)
summary(insurance_mod_race)

#Model with insurance and race*insurance
insurance_mod_raceint<-coxph(Surv(surv_mo, event_f)~insurance_f + race +race*insurance_f, cerv)
summary(insurance_mod_raceint)

#Test the hypothesis with the lrtest
lrtest(insurance_mod_race, insurance_mod_raceint)
```

# The interaction is significant, so the next step is to run models stratified by race
```{r}
#Whites
insurance_mod<-coxph(Surv(surv_mo, event_f)~insurance_f, cerv[which(cerv$race=="White"),])
summary(insurance_mod)

#Blacks
insurance_mod_black<-coxph(Surv(surv_mo, event_f)~insurance_f, cerv[which(cerv$race=="Black"),])
summary(insurance_mod_black)

#Hispanics
insurance_mod_hispanic<-coxph(Surv(surv_mo, event_f)~insurance_f, cerv[which(cerv$race=="Hispanic"),])
summary(insurance_mod_hispanic)

```

# Using contrasts to compare different combinations of race/ethnicity  and insurance
```{r}
#Let's answer the question-Is the hazard of death for those with Medicaid different between Blacks and Hispanics? We can use the model above starting on line 75 to answer this question using contrasts. 

#lets rerun the summary of the model
summary(insurance_mod_raceint)

#First we need to create a contrast vector that tests the null hypotheses of differences in the log hazard between these two groups (Blacks with Medicaid vs. Hispanics with Medicaid).  

#There are 8 values in our contrast vector, 1 for each coefficient in the model. 

#Let's determine the values to use in our contrast vector in excel--go to the excel sheet

#We will use these values and the coefficients to solve the equation below that calculates the difference in the log hazard of death between Blacks with Medicaid and Hispanics with Medicaid. 

#log_hazard= insurance_fAnyMedicaid*0 + insurance_fUninsured*0 + raceBlack*1 + raceHispanic*-1 + insurance_fAnyMedicaid:raceBlack*1 + insurance_fUninsured:raceBlack*0 + insurance_fAnyMedicaid:raceHispanic*-1 + insurance_fUninsured:raceHispanic*1 

#Then we will exponentiate the log hazard to get the hazard ratio of death in Blacks with Medicaid compared to Hispanics with Medicaid. 

#exp(insurance_fAnyMedicaid*0 + insurance_fUninsured*0 + raceBlack*1 + raceHispanic*-1 + insurance_fAnyMedicaid:raceBlack*1 + insurance_fUninsured:raceBlack*0 + insurance_fAnyMedicaid:raceHispanic*-1 + insurance_fUninsured:raceHispanic*1)

#Three different methods to solve the equation above on line 113 and 117 are below using R code are shown below.
```

# Method 1: plug in the coefficient values and constrast values into the formula (error prone and not reproducible if any changes to the dataset or model are made)
```{r}
log_hazard= (0.55756*0) + (0.72588*0) + (0.41809*1) + (-0.03708*-1) + (-0.24307*1) + (-0.25975*0) + (-0.24308*-1) + (-0.44640*0)
exp(log_hazard)
```

# Method 2: extract coefficient values and plug them and the contrast values into the formula (slightly less error prone)
```{r}
log_hazard= (summary(insurance_mod_raceint)$coefficients[1,1]*0) + (summary(insurance_mod_raceint)$coefficients[2,1]*0) + (summary(insurance_mod_raceint)$coefficients[3,1]*1) + (summary(insurance_mod_raceint)$coefficients[4,1]*-1) + (summary(insurance_mod_raceint)$coefficients[5,1]*1) + (summary(insurance_mod_raceint)$coefficients[6,1]*0) + (summary(insurance_mod_raceint)$coefficients[7,1]*-1) + (summary(insurance_mod_raceint)$coefficients[8,1]*0)
exp(log_hazard)
```

# Method 3: Using the multcomp package (even more less error prone)
```{r}
#There is a simpler way to do this creating a matrix of the values that will be multiplied by each coefficient and using that matrix with the general linear Hypothesis testing function (glht)
X<-matrix(c(0,  0, 1, -1, 1,  0, -1, 0),1)

t <- glht(insurance_mod_raceint, linfct = X) #linfct is the linear hypothesis to be tested as specified by X the matrix. In this case, we hypothesize that there is no difference in the hazard of death from cervical cancer between Blacks with Medicaid and Hispanics with Medicaid.
summary(t) #here are the results

#We can get the HR by exponentiating the coefficient
HRblack<-exp(0.4552)
HRblack

#We can also got 95% CI using the confint function
CI<-confint(t)
CI
lowCIblack<-exp(0.3431)
highCIblack<-exp(0.5673)
lowCIblack
highCIblack

#We can create a nice table with the HR and 95% CIs
table1<-data.frame(matrix(ncol=5, nrow=0)) #create the table
columns<-c("Race/Ethnicity", "Insurance", "HR",  "LowerCI", "UpperCI") #label the columns with names
colnames(table1)<-columns
#Plug in the data below
table1[1,]<-c('Hispanic',  'Medicaid', 1.0, "ref" , "ref")
table1[2,]<-c('Black',  'Medicaid', format(HRblack, digits=3), format(lowCIblack, digits=3), format(highCIblack, digits=3))
table1
```


```{r}
#For more information: https://stats.idre.ucla.edu/r/faq/how-can-i-test-contrasts-in-r/ 
```


